---
title: "ClimaticEffects-PA2"
author: "SNB"
date: "Sunday, July 25, 2015"
output: 
  html_document:
    keep_md: yes
---
#Effect of Storms on Costs of Casualties and Material Damage across the USA, 1950-2011.

##SYNOPSIS
This analysis is addressed to city, county or state planners, responsible for including health support and 
property protection measures in their budgets. It is a preliminary survey of the effect of the ten most significant 
climatic factors on human and material damage across the nation and over the 6 decades. 
An illustration is given of how the database may be used to extract more detailed and particular information about 
the particular locality or time zone for the relevant climatic dangers. 
An analysis of the trend over time in these parameters is beyond the scope of present report.

**References:** 

1)  Database:       [National Oceanic and Atomospheric Authority](https://d396qusza40orc.cloudfront.net/repdata%2fdata%2fstormdata.csv.bz2 )

2)  Documentations: [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf )

3)  FAQ:            [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC Storm Events-FAQ Page.pdf)


##DATA COLLECTION

Reading the data (for the second time after examining for inconsistencies:

```{r}
StormData<-read.csv(bzfile("repdata-data-StormData.csv.bz2","rt"),header=TRUE, na.strings=c("","+","?","-"), stringsAsFactors=FALSE)
```
**Summarising the data:**

```{r}
dim(StormData)
names(StormData)
head(StormData)
str(StormData)
summary(StormData)
summary(as.factor(StormData$TIME_ZONE))
summary(as.factor(StormData$EVTYPE))
summary(as.factor(StormData$PROPDMGEXP))
summary(as.factor(StormData$CROPDMGEXP))
summary(as.factor(StormData$TIME_ZONE))
```

##DATA PROCESSING

**DATA EXTRACTION**

Extracting the relevant data (10 columns taken from the original database) :
    The data BGN_DATA(column 2), time zone TIME_ZONE (column 4)state STATE (column 7), 
    event type EVTYPE (column 8), fatalities FATALITIES (column 23), injuries INJURIES (column 24),
    property damage PROPDMG (column 25), the corresponding dollar units PROPDMGEXP (column 26),
    crop damage CROPDMG (column 27) and the corresponding dollar units CROPDMGEXP (column 28)
    The LATITUDE and LONGITUDE data (columns 32,33) are also included, 
    in case these were more useful parameters than the time zone.
    
```{r}    
sdata <- StormData[,c(2,4,7,8,23,24,25,26,27,28,32,33)]
len   <-length(StormData$EVTYPE)

```
**INITIALISING with the required packets in R:**

```{r}
library(stringr)
library(ggplot2)
library(lattice)
library(dplyr)
```
**CLEANING up some observed anomalies in the EVTYPE data (no doubt many more to be found under "others"):**

```{r}

sdata$EVTYPE <- sapply(sdata$EVTYPE, function(x) toupper(x))
sdata$EVTYPE[sdata$EVTYPE=="AVALANCE"] <- "AVALANCHE"
sdata$EVTYPE[sdata$EVTYPE=="COASTALSTORM"] <- "COASTAL STORM"

```

###PROCESSING OF CASUALTY DATA

**PREPARING SUMMARIES OF CASUALTIES**

Forming tables of non-zero casualties
```{r}
casualties <- select(sdata, TIME_ZONE,EVTYPE, FATALITIES, INJURIES)
casualties <- filter(casualties, FATALITIES !=0 | INJURIES !=0)
lcas <-length(casualties$EVTYPE)

cat("There are a total of ",lcas, " non-zero records of casualties.")
```

**ARRANGING the death, injury and total casualty data by type of weather event and the time zone**

```{r}
LossAsCasualties <- group_by(casualties, EVTYPE, TIME_ZONE) %>%
                    summarize(Fatalities = sum(FATALITIES, na.rm=TRUE),
                              Injuries = sum(INJURIES, na.rm=TRUE),
                              TotalCasualties = Fatalities + Injuries)
```
**COMPRESSING the casualty data to the national plane:**
```{r}

LAC <-group_by(LossAsCasualties,EVTYPE)%>%
      summarise(National_Deaths= sum(Fatalities), 
                National_Injuries=sum(Injuries), 
                National_Casualties=sum(TotalCasualties))
LACd <-arrange(LAC,desc(National_Deaths))
LACi <-arrange(LAC,desc(National_Injuries))
LACt <-arrange(LAC,desc(National_Casualties))
Death_shortlist<-head(LACd,10)
Injuries_shortlist<-head(LACi,10)
Casualties_shortlist<-head(LACt,10)
                
```


##RESULTS FOR CASUALTIES on the national plane


The data on the human losses as deaths and injuries are produced in the following table and  barplots:

**National Statistics on Principle Causes of Deaths - 1950-2011**
```{r}
Death_shortlist
```
**National Statistics on Principle Causes of Injuries - 1950-2011**
```{r}
Injuries_shortlist
```
**National Statistics on Principle Causes of Caualties - 1950-2011**
```{r}
Casualties_shortlist

par(mfrow=c(1,3), mar = c(15,4,6,1), cex=0.75)

barplot(Death_shortlist$National_Deaths, las=3, names.arg = Death_shortlist$EVTYPE,
        ylab="All deaths",col="red")
barplot(Injuries_shortlist$National_Injuries, las=3, names.arg = Injuries_shortlist$EVTYPE,
        main="The ten main causes of \ndeaths and injuries \nacross USA - 1950-2011",
        ylab="All injuries", col="orange")
barplot(Casualties_shortlist$National_Casualties, las=3, names.arg = Casualties_shortlist$EVTYPE,
        ylab="All casualties", col="black")
        
```


###PROCESSING OF MATERIAL LOSS DATA

**PREPARING SUMMARIES OF MATERIAL DAMAGE (property and crop damage)**

Forming tables of non-zero damage
```{r}
damages <- select(sdata, TIME_ZONE,EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
damages <- filter(damages, PROPDMG !=0 | CROPDMG !=0)
damages$PROPDMGEXP <- sapply(damages$PROPDMGEXP, function(x) toupper(x))
damages$CROPDMGEXP <- sapply(damages$CROPDMGEXP, function(x) toupper(x))
ldam <-length(damages$EVTYPE)

cat("There are a total of ", ldam, "non-zero records of material costs")
```

**CONVERTING costs to the common unit, million dollars.**
```{r}
summary(as.factor(damages$PROPDMGEXP))
summary(as.factor(damages$CROPDMGEXP))

damages$PROPDMGEXP[damages$PROPDMGEXP=="H"] <- "2"
damages$PROPDMGEXP[damages$PROPDMGEXP=="K"] <- "3"
damages$PROPDMGEXP[damages$PROPDMGEXP=="M"] <- "6"
damages$PROPDMGEXP[damages$PROPDMGEXP=="B"] <- "9"
damages$CROPDMGEXP[damages$CROPDMGEXP=="K"] <- "3"
damages$CROPDMGEXP[damages$CROPDMGEXP=="M"] <- "6"
damages$CROPDMGEXP[damages$CROPDMGEXP=="B"] <- "9"

summary(as.factor(damages$PROPDMGEXP))
summary(as.factor(damages$CROPDMGEXP))

damages$PROPDMGEXP <- as.integer(damages$PROPDMGEXP)
damages$CROPDMGEXP <- as.integer(damages$CROPDMGEXP)

damages$PROPDMG <- damages$PROPDMG * 10**(damages$PROPDMGEXP-6)
damages$CROPDMG <- damages$CROPDMG * 10**(damages$CROPDMGEXP-6)

summary(damages$PROPDMG)
summary(damages$CROPDMG)

```
**ARRANGING the property, crop and total material loss data by type of weather event and the time zone**

```{r}
LossAsDamages <- group_by(damages, EVTYPE, TIME_ZONE) %>%
                    summarize(PropertyDamages = sum(PROPDMG, na.rm=TRUE),
                              CropDamages = sum(CROPDMG, na.rm=TRUE),
                              TotalDamages = PropertyDamages + CropDamages)
```
**COMPRESSING the damages data to the national plane:**
```{r}

LAD <-group_by(LossAsDamages,EVTYPE)%>%
      summarise(Property_Loss = sum(PropertyDamages), 
                Crop_Loss=sum(CropDamages), 
                Damages=sum(TotalDamages))
LADP <-arrange(LAD,desc(Property_Loss))
LADC <-arrange(LAD,desc(Crop_Loss))
LADT <-arrange(LAD,desc(Damages))
Property_shortlist<-head(LADP,10)
Crop_shortlist<-head(LADC,10)
Damages_shortlist<-head(LADT,10)
                
```

##RESULTS FOR MATERIAL LOSSES on the national plane

The data on the material damages as property and crops losses are produced in the following table and  barplots:

**National Statistics on Principle Causes of Property Damages - 1950-2011**
```{r}
Property_shortlist
```
**National Statistics on Principle Causes of Crop Damages - 1950-2011**
```{r}
Crop_shortlist
```
**National Statistics on Principle Causes of Material Losses - 1950-2011**
```{r}
Damages_shortlist

par(mfrow=c(1,3), mar = c(15,4,6,1), cex=0.75)

barplot(Property_shortlist$Property_Loss, las=3, names.arg = Property_shortlist$EVTYPE,
        ylab="All property losses", col="blue")
barplot(Crop_shortlist$Crop_Loss, las=3, names.arg = Crop_shortlist$EVTYPE,
        main="The ten main factors \naffecting economic loss \nacross USA - 1950-2011 \n(million dollars)",
        ylab="All rop losses",col="green")
barplot(Damages_shortlist$Damages, las=3, names.arg = Damages_shortlist$EVTYPE,
        ylab="All material losses",col="brown")
        
```


##PROCESSING of data for the MAJOR CLIMATIC FACTORS affecting casualities and damages for selected TIME ZONES

###Investigating the patterns over time zones for the six main climatic events

```{r}
casualties_timezone<-filter(LossAsCasualties,EVTYPE=="TYPHOON"|EVTYPE== "FLOOD"|EVTYPE=="TORNADO"|EVTYPE=="DROUGHT"|EVTYPE=="EXCESSIVE HEAT")
damages_timezone   <-filter(LossAsDamages,EVTYPE=="TYPHOON"|EVTYPE== "FLOOD"|EVTYPE=="TORNADO"|EVTYPE=="DROUGHT"|EVTYPE=="EXCESSIVE HEAT")
head(casualties_timezone, 40)
summary(as.factor(casualties_timezone$TIME_ZONE))
head(damages_timezone, 40)
summary(as.factor(damages_timezone$TIME_ZONE))
```

###Selecting the maximum number of time zones to give enough data for comparison

```{r}
casualties_4zones<-filter(casualties_timezone,TIME_ZONE=="PST"|TIME_ZONE=="CST"|TIME_ZONE=="MST"|TIME_ZONE=="EST")
damages_3zones<-filter(damages_timezone,TIME_ZONE=="PST"|TIME_ZONE=="CST"|TIME_ZONE=="EST")

head(casualties_4zones,20)
head(damages_3zones,20)
casualties_4zones <- casualties_4zones[2:13,]
damages_3zones <- damages_3zones[1:12,]
```

##RESULTS for the selected major climatic events for a few time zones

```{r}
head(casualties_4zones,20)
head(damages_3zones,20)

par(mar=c(8,3,8,1))
damages_3zones$TIME_ZONE    <- as.factor(damages_3zones$TIME_ZONE)

barchart(TotalDamages~EVTYPE | TIME_ZONE, damages_3zones,layout=c(2,2),col = c("orange","red","blue","black"),
                                main="Example: Damage across 3 time zones for 4 strong climatic events",
                                scales=list(x=list(cex=0.45)))

```


##CONCLUSION

As can be seen, by far the most prominent cause of death and injuries as registered over the whole of USA 
between 1959 and 2011 is associated with tornados, while other factors such as flood and heat related events are of the same 
order of magnitude. 

On the other hand in the same period across the nation, floods have almost doubled the economic damage to property 
due to the other important factors, such as typhoon and tornados, and the pattern is the same for the overall material 
damage. 

In the case of crops, economic loss due to drought is double that due to floods and the other relevant climatic events, 
but crop losses make up a smaller portion of the entire material damage costs.

As is expected, the pattern of losses measured in human or economic terms depends on the time zone (or, more exactly, on the
latitude and longitude) of the observations and varies appreciably from the national pattern.

Whether the patterns over time are influenced by a real trend or by exceptional weather and climate effects can only
be decided after grouping the data by year or decade, and is beyong the scope of this report.

#http://rpubs.com/SNB/97741